---
title: "20230721_RNAseq_SL1344"
author: "Hanna Fricke"
date: "2023-07-21"
output: html_document
---
The intra macrophage experiment was conducted by Canals et al in their 2019 paper. They used the ST74 strain in three replicates in different in vitro conditions as well as intramacrophage growth. Here I used their raw count data they provide for their DE analysis in DEGUST

paper:
https://doi.org/10.1371/journal.pbio.3000059

# Expression analysis of ST74 macrophage model
## set up working directory

```{r}
# set result directory for plots etc.
data_dir <- "/home/hanna/Documents/Macrophages_project/Final/M_RNAseq/data/"
res_dir <- "/home/hanna/Documents/Macrophages_project/Final/M_RNAseq/output/"
excel<-"/home/hanna/Documents/Macrophages_project/Final/M_RNAseq/output/excel/"
images<-"/home/hanna/Documents/Macrophages_project/Final/M_RNAseq/output/images/"

if(!dir.exists(res_dir)) dir.create(res_dir)
if(!dir.exists(excel)) dir.create(excel)
if(!dir.exists(images)) dir.create(images)
knitr::opts_chunk$set(fig.path=res_dir, warning=F,
                      message=F, echo=T, dev="cairo_pdf")

if(!dir.exists(res_dir)) dir.create(res_dir)
```

## Load some packages

```{r}
library(matrixStats)
library(magrittr)
library(ggplot2)
```

## Load data and create count table

```{r}
library(data.table)
library(SummarizedExperiment)
## load and format files
# macro data
macro_all<-fread(paste0(data_dir,"Rawseq_reads_fitness_strain.csv"),data.table = F) # This file contains CDS and sRNA data from Wang et al
ST74<-macro_all[,c(2:3,4:6,16:18)] # select ESP and macro
rownames(ST74)<-ST74$`FEATURE`
meta_data<-ST74[,c(1:2)] # get meta data
colnames(meta_data)<-c("locus_tag","gene_name") # assign appropriate col names

colnames(ST74)<-c("locus_tag","gene_name","ESP_1","ESP_2","ESP_3","Macro_1","Macro_2","Macro_3")

counts_tab<-ST74[,-c(1:2)]%>%data.matrix() # remove gene name and locus 

```


# EdgeR -Preprocessing

## Initialise edgeR variables

```{r}
library("limma")
library('edgeR')
## Get metadata
sample_info<-data.frame(identifier=c("ESP_1","ESP_2","ESP_3","Macro_1","Macro_2","Macro_3"), # "input1",
                        condition=c(rep("ESP",3),rep("Macro",3)),
                        replicate=c("1","2","3","1", "2","3"))
group <- factor(sample_info$condition) # condition grouping factor
replicate <- factor(sample_info$replicate) # replicate factor

dgeAll <- DGEList(counts_tab, group=group, genes=rownames(counts_tab)) # make a DGE list with group indicator=column and rows= feature
```

### PCA of raw count data
PCA Raw counts 
```{r}
library('viridis')
library('ggplot2')
logcounts<-log2(counts_tab+1)
fname<-"RNAseq_"
pdf(paste0(images, "PCA_Macro_raw_log2counts.pdf"))
pca_raw <- stats::prcomp(t(logcounts))  # :: helps you to access specific package in r (here stats:) then apply pcrcomps to transposed logcounts
pca_df <- cbind(pca_raw$x %>% as.data.frame(),   # make a PCA df by cbind (column bind), make it a df with x=PCA components  and add a column with sample_names
sample=rownames(pca_raw$x))
pca_df$condition <- sample_info$condition[match(pca_df$sample, sample_info$identifier)]# assign meta data to the sample

pca_imp <- summary(pca_raw)$importance[2,] # importance = variance explained
# plot the PCA1 and PCA2
p <- ggplot(pca_df, aes(x=PC1, y=PC2,label=replicate, col=factor(condition))) +
  geom_point(aes(shape=replicate),size=4.5)+
  scale_color_manual("condition", values=viridis(3)) +
  scale_size_manual("", values=c(7,4)) +
  scale_x_continuous(name=paste0("PC1 (", round(pca_imp[1]*100), "%)")) +
  scale_y_continuous(name=paste0("PC2 (", round(pca_imp[2]*100), "%)")) +
  theme_classic() +
  theme(axis.text=element_text(size=20,color="black"),
        axis.title=element_text(size=23,color="black",face="bold"),
        plot.title=element_text(size=23,face="bold"),
        legend.text=element_text(size=21),
        legend.title=element_text(size=23,face="bold"),
        legend.spacing.x=unit(0.1, "cm"),
        legend.position="top",
        legend.box="vertical",
        legend.margin=margin())
print(p)
dev.off()
print(p)
```


## Filter based on cpm to remove low counts

```{r}
# rule of thumb to identify the ideal threshold
edgeR::cpm(10, mean(dgeAll$samples$lib.size)) # 3 cpm ~ 10 counts per lib
# Remove all counts that are too low 
keep <- rowSums(edgeR::cpm(dgeAll$counts) >= 3)>=4# filter based on input in at least two replicates
dgeAll <- dgeAll[keep, , keep.lib.sizes=F] # keep.lib.sizes=FALSE, the lib.size for each sample will be recalculated 
# inspect the filtering result
dim(dgeAll$counts)  # to check how many rows were removed
```


## Normalise data using edgeRs TMM method

```{r,echo=FALSE, fig.show='hide'}

dgeAll  <- calcNormFactors(dgeAll,logratioTrim=0.35, sumTrim=0.08) # Increased logTrim due to spread of data rather than clustering of individual replicates around 0

#calculate the normalised counts 
normCounts <- edgeR::cpm(dgeAll) # divide by library size and multiply by 1mio
logNormCounts <- log2(normCounts + 1)
```

### PCA of TMM normalised counts

```{r}
pdf(paste0(images, "PCA_Macro_norm_counts.pdf"))
logcounts <- logNormCounts 
pca_raw <- stats::prcomp(t(logcounts))  # :: helps you to access specific package in r (here stats:) then apply pcrcomps to transposed logcounts
pca_df <- cbind(pca_raw$x %>% as.data.frame(),   # make a PCA df by cbind (column bind), make it a df with x=PCA components  and add a column with sample_names
sample=rownames(pca_raw$x))
pca_df$condition <- sample_info$condition[match(pca_df$sample, sample_info$identifier)]# assign meta data to the sample

pca_imp <- summary(pca_raw)$importance[2,] # importance = variance explained
# plot the PCA1 and PCA2
p <- ggplot(pca_df, aes(x=PC1, y=PC2,label=replicate, col=factor(condition))) +
  geom_point(size=4.5)+
  scale_color_manual("condition", values=viridis(3)) +
  scale_size_manual("", values=c(7,4)) +
  scale_x_continuous(name=paste0("PC1 (", round(pca_imp[1]*100), "%)")) +
  scale_y_continuous(name=paste0("PC2 (", round(pca_imp[2]*100), "%)")) +
  theme_classic() +
  theme(axis.text=element_text(size=20),
        axis.title=element_text(size=23),
        plot.title=element_text(size=23,face="bold"),
        legend.text=element_text(size=21),
        legend.title=element_text(size=23,face="bold"),
        legend.spacing.x=unit(0.1, "cm"),
        legend.position="top",
        legend.box="vertical",
        legend.margin=margin())
print(p)
dev.off()
print(p)
```

# EdgeR- Differential fitness analysis

## Set up design matrix
```{r, echo=FALSE}
designMat <- model.matrix(~0+group)# model without intercept
colnames(designMat) <- colnames(designMat) %>% sub("group", "", .) # remove group from groupOUTPUT
contrast<- makeContrasts(Macro-ESP, levels=designMat) 
```

## Estimate dispersion

```{r, echo=FALSE}
dgList <- estimateDisp(dgeAll, design=designMat) 
plotBCV(dgList) 
```

## Fit NB model and F-test

```{r}
##Fit the model
fit <- glmQLFit(dgList, designMat, robust=TRUE) # Fit a quasi-likelihood negative binomial glm, conduct genewise statistical tests 
plotQLDisp(fit)

## conduct an F-test based multiple comparison test
qlf<- glmQLFTest(fit, contrast=contrast) # empirical Bayes quasi-likelihood F-tests > stricter error rate control as it accounts for uncertainty in dispersion
```

### Inspect the results and format

```{r}
# Extract the top DE genes 
result2<- topTags(qlf, length(logNormCounts[,1])) # get all genes
# get a df with all the results and add relevant information
result2<-result2$table # DE results

```

### Make a dataframe for plotting and further inspection

```{r}
# Quick assignment, not thorough assignment of gene names and function
my_data<-fread(paste0(data_dir,"TraDIS_DE_Pl123.csv"),data.table=F)#for spi assignment
names(result2)[names(result2)=="genes"]<-"locus_tag"
result2$gene_name<-meta_data$gene_name[match(result2$locus_tag, meta_data$locus_tag)] #add gene name
result2$gene_function<-meta_data$gene_function[match(result2$locus_tag, meta_data$locus_tag)] #add function
result2$spi<-my_data$spi[match(result2$gene_name,my_data$gene_name)]
```

### Density plot of genes with and without fitness effect

```{r}
library(ggridges)
library(plyr)
resultDE_1<-result2
resultDE_1$model<-""
resultDE_1$DE2<-ifelse(resultDE_1$FDR<0.001, "Yes", "No")

pdf(paste0(images,"Density_distributions_DF_effects_MAcro-edgeronly<FDR0.001.pdf"), width = 15, height =12)
hela_edger<-ggplot(resultDE_1, aes(x=logFC, y=model,group = interaction(DE2, model), fill = DE2)) +
  geom_density_ridges(alpha=0.4) +
  theme_ridges() + 
  theme(axis.text.y=element_text(size=30,color="black"),
        axis.title.y = element_blank(),
        axis.text.x=element_text(size=30,color="black"),
        axis.title.x=element_text(size=30,color="black",hjust=0.5, vjust=0.9,face="bold"),
        plot.title=element_blank(),
        legend.text=element_text(size=30),
        legend.title=element_text(size=30,face="bold"),
        legend.spacing.x=unit(0.1, "cm"),
        legend.position="right",
        legend.justification = c("left","top"),
        legend.box="vertical",
        legend.margin=margin(),
        axis.line = element_line(color = "black"))+
  guides(fill=guide_legend(title="Fitness effect:"))+
  xlab("Log2FC (Output/Input)")+
  scale_fill_manual("DE2",values=c("darkgrey", "#31688EFF"))+
  geom_vline(xintercept=c(-1.25,0,1.25),linetype="dashed",linewidth=2, color = c("#31688EFF", "lightgrey","#31688EFF"))+
    scale_x_continuous(breaks=seq(round_any(min(resultDE_1$logFC),10), round_any(max(resultDE_1$logFC),10),1)) +
  scale_y_discrete(expand = c(0, 0))

print(hela_edger)
dev.off()
print(hela_edger)

# resultDE_1$DE2<-ifelse(resultDE_1$FDR<0.05, "Yes", "No")
# pdf(paste0(images,"Density_distributions_DF_effects_MAcro-edgeronly<FDR0.05.pdf"), width = 20, height=18)
# hela_edger<-ggplot(resultDE_1, aes(x=logFC, y=model,group = interaction(DE2, model), fill = DE2)) +
#   geom_density_ridges(alpha=0.4) +
#   theme_ridges() + 
#   theme(axis.text.y=element_text(size=30,color="black"),
#         axis.title.y = element_blank(),
#         axis.text.x=element_text(size=30,color="black"),
#         axis.title.x=element_text(size=30,color="black",hjust=0.5, vjust=0.9,face="bold"),
#         plot.title=element_blank(),
#         legend.text=element_text(size=30),
#         legend.title=element_text(size=30,face="bold"),
#         legend.spacing.x=unit(0.1, "cm"),
#         legend.position="right",
#         legend.justification = c("left","top"),
#         legend.box="vertical",
#         legend.margin=margin(),
#         axis.line = element_line(color = "black"))+
#   guides(fill=guide_legend(title="Fitness effect:"))+
#   xlab("Log2FC (Output/Input)")+
#   scale_fill_manual("DE2",values=c("darkgrey", "#31688EFF"))+
#   geom_vline(xintercept=c(-1.25,0,1.25),linetype="dashed",linewidth=2, color = c("#31688EFF", "lightgrey","#31688EFF"))+
#     scale_x_continuous(breaks=seq(round_any(min(resultDE_1$logFC),10), round_any(max(resultDE_1$logFC),10),1)) +
#   scale_y_discrete(expand = c(0, 0))
# print(hela_edger)
# dev.off()
# print(hela_edger)
```
 Higher FDR results in clearer distinction. Cut off at 1.25
 
### MDA plot

```{r}
# 
#assign fitness category
result2$DE<-ifelse(result2$FDR< 0.001& result2$logFC<=-1.25,"Attenuated",ifelse(result2$FDR<0.001& result2$logFC >=1.25,"Amplified", "Unchanged"))

## MD plot
pl2 <- ggplot(result2) +
    geom_point(data=result2,aes(x=logCPM,y=logFC))
print(pl2)
dev.off()             
pl2 <- ggplot(result2) +
    geom_point(data=result2[which(result2$DE == "Unchanged"),],aes(x=logCPM,y=logFC,colour='Unchanged'),shape=19,  size=5,alpha = 1/3)+
  geom_point(data=result2[which(result2$DE == "Amplified"),],aes(x=logCPM,y=logFC,colour='Amplified'),  shape=19, size=5,alpha=0.5)+ 
  geom_point(data=result2[which(result2$DE == "Attenuated"),],aes(x=logCPM,y=logFC,colour='Attenuated'), shape=19,size=5, alpha=0.5) +
  geom_hline(yintercept=c(0,1,-1),linetype="dashed", color = c("black", "lightgrey","lightgrey"))+
   geom_point(data=result2[which(result2$FDR>0.05 & result2$spi== "SPI-2"),],aes(x=logCPM,y=logFC,colour='SPI-2'),alpha = 2/3,size=5)+
    geom_point(data=result2[which(result2$FDR<0.05 &result2$spi == "SPI-2"),],aes(x=logCPM,y=logFC,colour='SPI-2'),alpha = 1,size=4.5)+
    geom_point(data=result2[which(result2$FDR>0.05 & result2$spi== "SPI-1"),],aes(x=logCPM,y=logFC,colour='SPI-1'),alpha = 2/3,size=5)+
    geom_point(data=result2[which(result2$FDR<0.5& result2$spi == "SPI-1"),],aes(x=logCPM,y=logFC,colour='SPI-1'),alpha = 1,size=4.5)+
  scale_color_manual("", values=c("Amplified"="#38598CFF", 
                                  "Attenuated"="#440154FF",
                                  "Unchanged"="lightgrey",
                                  "SPI-1"= "#35B779FF", 
                                  "SPI-2"= "#FFC725CC"),#"#35B779FF" ,"#FDE725FF"
                     labels= ~ stringr::str_wrap(.x, width = 1)) +
  scale_size_manual("Log2FC Fitness", values=c(7,4)) +
  scale_x_continuous("Average expression log2(cpm)") +
  scale_y_continuous("Log2FC (Output/Input)") +
  theme_classic() +
  theme(axis.text=element_text(size=30,color="black"),
        axis.title=element_text(size=30,color="black",face="bold"),
        legend.text=element_text(size=20),
        legend.title=element_text(size=30,face="bold"),
        legend.spacing.x=unit(0.1, "cm"),
        legend.position="right",
        legend.justification=c("left","top"))
print(pl2)
cairo_pdf(paste0(images, "TraDIS_MDA_edgeRonly_FDR<0.001&logFC1.25.pdf"), width = 12, height = 10) # have issues with paste so will jsut specify the whole pathway name
print(pl2)
dev.off()
png(paste0(images, "TraDIS_MDA_macro_expr_edgeRonly_FDR<0.001&logFC1.25.png"), width = 4000, height = 4000, res=300)
print(pl2)
dev.off()

print(pl2)
```



### Vulcano plot 

```{r}
library(ggrepel)
plot_data<-result2
pdf(paste0(images,"Figure_DE_vulcano_Macro_expr_edgeR_FDR<0.001&logFC1.25.pdf"), width = 25, height = 20)
  vulcano<-ggplot(data =plot_data , aes(x = logFC, y = -log10(FDR),col = DE)) +
    geom_point(alpha=0.5,size=3)+
     geom_point(data=result2[which(grepl("hfq",result2$gene_name)==T|grepl("pyr",result2$gene_name)==T|grepl("arg",result2$gene_name)==T|grepl("pur",result2$gene_name)==T),],aes(x=logFC,y=-log10(FDR),colour='Pur genes'),alpha = 1,size=6.5)+
     geom_point(data=result2[which(result2$FDR>=0.001 & result2$spi== "SPI-2"),],aes(x=logFC,y=-log10(FDR),colour='SPI-2'),alpha = 0.5,size=5)+
    geom_point(data=result2[which(result2$FDR<0.001 &result2$spi == "SPI-2"),],aes(x=logFC,y=-log10(FDR),colour='SPI-2'),alpha = 1,size=4.5)+
    geom_point(data=result2[which(result2$FDR>=0.001  & result2$spi== "SPI-1"),],aes(x=logFC,y=-log10(FDR),colour='SPI-1'),alpha = 0.5,size=5)+
    geom_point(data=result2[which(result2$FDR<0.001& result2$spi == "SPI-1"),],aes(x=logFC,y=-log10(FDR),colour='SPI-1'),alpha = 1,size=4.5)+
    theme_set(theme_classic(base_size = 20) +
              theme(
                axis.title.y = element_text(hjust = 0.5,face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
                axis.title.x = element_text(hjust = 0.5, margin = margin(0,20,0,0), size = rel(1.1),face = "bold", color = 'black'),
                plot.title = element_text(hjust = 1),
                legend.title=element_text(size=20,face="bold",color="black"),
                legend.text=element_text(size=20,face="bold",color="black"),
                legend.position="right",
                legend.justification = c("left", "top")))+
                geom_vline(xintercept = c(-1.25, 1.25), col = "gray", linetype = 'dashed') +
                geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
                scale_color_manual(values = c("Attenuated"="#440154FF", 
                                              "Amplified"="#38598CFF",
                                              "Unchanged"="grey", 
                                              "SPI-1"= "#35B779FF",
                                              "SPI-2"= "#FFC725CC",
                                              "Pur genes"= "red"),breaks=c("Attenuated", "Amplified","Unchanged", "SPI-1","SPI-2")) + # to set the colours of our variabe
                guides(color=guide_legend(title=""))+
        geom_text_repel(label=ifelse( grepl("pyr",plot_data$gene_name)==T|grepl("pur",plot_data$gene_name)==T|grepl("arg",result2$gene_name)==T|grepl("hfq",plot_data$gene_name)==T,plot_data$gene_name,''),box.padding = 0.3,max.overlaps = Inf,position = position_jitter(0.1, seed = 2), show.legend = FALSE,col="black",fontface = "bold", size=15)
  print(vulcano)
  dev.off()
```

### heatmap 

```{r}
result2DE<-result2[result2$FDR<0.001&abs(result2$logFC)>=1.25,]
fname=paste0(images,"Macro_heatmap_edgeR_FDR<0.001&logFC1.25")
  # Add the gene names to the data
  # order the top values and get the 
  logCPM <- data.frame(logNormCounts) # log transform values and take cpm , +2 to avoid undefined log2(0)
  logCPM$FDR<-result2$FDR[match(rownames(logCPM),rownames(result2))]
  o <- order(logCPM$FDR) # returns indices of P-values in sorted manner
  logCPM <-logCPM[o[1:length(result2DE[,1])],] # plot top genes --> select desired range
  logCPM<-logCPM[,c(1:length(logCPM[1,])-1)] # remove FDR ag
  logCPM <- t(scale(t(logCPM))) # scale to mean 0 and sd 1
  
  # Create the heatmap annotation
  # column annotations
  annotation_col = data.frame(
                      Condition = factor(c(rep(c("Input"),3),rep(c("Output"), 3))))
  rownames(annotation_col)<-colnames(logCPM)
  annotation_row = data.frame(Genes=rownames(logCPM))
  #row annoations
  
  rownames(annotation_row)<-annotation_row$Genes
  annotation_row$logFC<-result2$logFC[match(rownames(logCPM),rownames(result2))]
  annotation_row<-annotation_row$logFC %>% as.data.frame(row.names = rownames(annotation_row))
  colnames(annotation_row)="LogFC"
  #colour annotations
  ann_colors = list(
    Condition = c(Input = "grey", Output = "#000004FF"),
    LogFC=colorRampPalette(c("navy","#FFFFBF", "#A50026"))(50))
  heatmap<-ComplexHeatmap::pheatmap(logCPM, 
           cluster_rows = T,                       
           cluster_cols = T, 
           show_rownames = F, 
           show_colnames = T,
           treeheight_row= 0,
           cutree_rows = 2,
           cutree_cols = 2,
           annotation_col = annotation_col, 
           annotation_row = annotation_row,
           annotation_names_row = F,
           annotation_names_col = F,
           name = "Log2 (cpm)",
           annotation_colors = ann_colors,
           col=colorRampPalette(c( "#38598CFF","#F7F7F7","#440154FF"))(50),
           clustering_distance_rows = "euclidean")
  print(heatmap)
  cairo_pdf(paste0(fname,".pdf") , width = 6, height = 6) # have issues with paste so will jsut specify the whole pathway name CHANGE name for pseudo/no pseudo gene normalisation
  print(heatmap)
  dev.off()
  png(paste0(fname, ".png"), width = 2000, height = 2000, res=300)
  print(heatmap)
  dev.off()
  print(heatmap)
```

```{r}
write.csv(result2,paste0(excel,"RNA_seq_ST74.csv"), row.names = F)
result2DE<-result2[result2$FDR<0.001&abs(result2$logFC)>=1.25,]
```

```{r}
sessionInfo()
```

